---
title: "REVIMAR-MAP: Benthic Marine Habitats Mapping"
author: "Tiago Gandra"
date: today
format:
  html:
    toc: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Project Overview

The **REVIMAR-MAP** project aims to provide a cartographic and methodological baseline for the predictive classification of benthic marine habitats on the Brazilian continental margin. This open-science workflow generates wall-to-wall spatial models to support Marine Spatial Planning (MSP) and environmental assessments, aligning with the guidelines of the Ministry of Environment and Climate Change (MMA) and actions funded by FUNBIO.

### 1.1. Scripts and Data

To ensure full transparency and reproducibility, this project separates the source code from the heavy geospatial outputs:
1. **L1: Geomorphological Provinces** (e.g., Shelf, Slope, Basin, Seamounts)
2. **L2: Biological Zones** (e.g., Photic, Mesophotic, Bathyal, Abyssal)
3. **L3: Substrate** (e.g., Mud, Sand, Gravel, Reefs)

* **Source Code:** The complete R scripts and processing pipelines are publicly available on GitHub at [https://github.com/gandrat/revimar](https://github.com/gandrat/revimar).
* **Geospatial Data:** Due to storage limitations for version control, the final models and their associated metadata are hosted and versioned on Zenodo. The datasets can be accessed via the following record: [https://zenodo.org/records/22834505](https://zenodo.org/records/22834505).

## 2. Level 1 - Oceanic Provinces

The methodology adopts a multiscalar top-down approach. The first level of the classification hierarchy (Level 1) partitions the marine environment into broad geomorphological provinces based on bathymetry, slope, and major structural features.

### 2.1. Data Preparation and Bathymetric Smoothing

All spatial data is standardized to the Albers Equal Area Conic projection (Brazil). The raw bathymetric data (e.g., GEBCO) is smoothed using a 7x7 focal mean filter to remove high-frequency noise and local artifacts before deriving morphometric variables.

```{r smoothing, eval=FALSE}
# Define Albers projection for Brazil
crs_albers_brasil <- "+proj=aea +lat_0=-12 +lon_0=-54 +lat_1=-2 +lat_2=-22 +x_0=5000000 +y_0=10000000 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"

# Load and project bathymetry
bat_gebco <- rast('data/raw/gebco/gebco_raw.tif')
bat <- project(bat_gebco, crs_albers_brasil)

# Clamp values to isolate marine areas (depths <= 0)
bat <- clamp(bat, upper = 0)

# Apply a 7x7 focal mean filter to smooth the terrain
bat <- focal(bat, w = 7, fun = "mean", na.rm = TRUE)
```

### 2.2. Slope Calculation

Slope is essential to distinguish flat plains (basins and shelves) from steep structural zones (slopes). A focal maximum filter is also applied to emphasize the steepest gradients.

```{r slope_calculation, eval=FALSE}
# Calculate slope in degrees
slope_deg <- terrain(bat, v = "slope", unit = "degrees")

# Apply focal max to highlight structural breaks
slope_degf <- focal(slope_deg, w = 7, fun = 'max', na.rm = TRUE)
```

### 2.3. Sequential Map Algebra (Province Classification)

We apply strict depth and slope thresholds using sequential map algebra to delineate the oceanic provinces.

```{r map_algebra, eval=FALSE}
# Create an empty raster template based on bathymetry
benthic_zones <- rast(bat)
values(benthic_zones) <- NA

# Class 4: Oceanic Basin (Abyssal Plains)
# Everything deeper than 5000m, regardless of minor slope variations
benthic_zones[is.na(benthic_zones) & bat <= -5000] <- 4

# Class 1: Continental Shelf
# Shallower than the shelf break (-250m) with gentle slopes, or strictly shallower than -90m
benthic_zones[bat >= -250 & slope_deg < 0.5] <- 1
benthic_zones[bat >= -90] <- 1

# Class 2: Continental Slope
# Intermediate depths with a steep gradient
benthic_zones[is.na(benthic_zones) & bat >= -2500] <- 2
benthic_zones[is.na(benthic_zones) & slope_deg >= 0.5 & bat >= -3500] <- 2

# Class 3: Continental Rise
# Remaining unclassified intermediate to deep waters with a gentle gradient
benthic_zones[is.na(benthic_zones) & slope_deg >= 0.1] <- 3
benthic_zones[is.na(benthic_zones) & bat <= -3500] <- 4
```

### 2.4. Seamounts Integration

Isolated topographic peaks (seamounts) mapped independently in QGIS are integrated into the classification as a distinct Level 1 feature.

```{r seamounts, eval=FALSE}
# Load seamounts vector and rasterize
seamounts_vec <- read_sf('data/processed/shapes/gis_revimar.gpkg', layer = 'seamounts')
seamounts_vec <- st_transform(seamounts_vec, crs_albers_brasil)

seamounts_rast <- rasterize(seamounts_vec, benthic_zones, field = 5, background = NA)

# Overlay seamounts onto the benthic zones
benthic_zones <- cover(seamounts_rast, benthic_zones)
```

### 2.5. Noise Reduction and Polishing

To create a cartographically sound product, we apply noise reduction algorithms. A focal modal filter smooths boundaries, and a sieve algorithm removes isolated pixels (salt-and-pepper effect). Seamounts are reapplied at the end to ensure small, valid features are not deleted by the sieve process.

```{r noise_reduction, eval=FALSE}
# 1. Apply focal modal filter for boundary smoothing
focal_window <- matrix(1, nrow = 21, ncol = 21)
benthic_zones_m <- focal(benthic_zones, w = focal_window, fun = 'modal', na.rm = TRUE)

# 2. Sieve: Remove small isolated polygon fragments (< 50000 pixels)
pixels_tresholds <- 50000
benthic_zones_c <- terra::sieve(benthic_zones_m, threshold = pixels_tresholds, directions = 8)

# 3. Restore seamounts to prevent accidental deletion by the sieve
benthic_zones_final <- cover(seamounts_rast, benthic_zones_c)
```
## 3. Level 2 - Biological Zones

The second hierarchical level of the REVIMAR-MAP project integrates biological proxies into the geomorphological base. Using climatological data from the Copernicus Marine Service (CMEMS), we model the benthic light availability (Euphotic vs. Aphotic zones) and stratify deep-water features into Bathyal and Abyssal zones.

### 3.1. Temporal Aggregation of Euphotic Depth (Zeu)

We use the CMEMS Euphotic Depth (`Zeu`) NetCDF product. To establish a robust climatological baseline, we calculate the cell-wise median across the time series and align it spatially with the Level 1 Geomorphological Provinces.

```{r zeu_aggregation, eval=FALSE}
# Load the L1 Provinces and Bathymetry
l1 <- rast('outputs/l1_benthic_provinces.tif')
bat <- rast('data/processed/bathymetry_gebco.tif')

# Load the Copernicus CMEMS Euphotic Depth (Zeu) NetCDF file
zeu <- rast('data/raw/cmems_zeu/cmems_mod_glo_bgc_my_0.083deg-lmtl_P1D-i_1773341382551.nc')

# Calculate the climatological median and match spatial extents
zeu_med <- median(zeu, na.rm = TRUE)
zeu_med <- project(zeu_med, l1)
zeu_med <- mask(zeu_med, bat)

```

### 3.2. Defining the Euphotic Benthos
Sunlight reaches the seafloor where the euphotic depth (Zeu) is greater than or equal to the water column depth (bathymetry). We create a boolean mask to map these productive shallow-water habitats.

```{r light_reaches, eval=FALSE}
# Create a boolean mask: TRUE where light reaches the bottom
zeu_bat <- zeu_med >= (bat * -1)

# Convert TRUE/FALSE to numeric 1 (Euphotic) and 0 (Aphotic)
zeu_bat_num <- ifel(zeu_bat, 1, 0)

```
### 3.3. Habitat Overlap and Deep Water Stratification
The physical provinces (Shelf, Slope, Rise, Basin, Seamounts) are intersected with the photic data. Furthermore, deep-water environments are strictly stratified into Bathyal (> -4000m) and Abyssal (<= -4000m) zones, following global marine ecology standards.

```{r reclass, eval=FALSE}
# Combine rasters into unique two-digit IDs: L1 (1-5) * 10 + Photic (0-1)
geom_photic <- (l1 * 10) + zeu_bat_num

# Clean-up: Deep environments (Slope, Rise, Basin) are inherently aphotic
# Stratification Threshold: 4000m depth boundary for Bathyal vs Abyssal
geom_photic <- ifel(geom_photic == 20 & bat <= -4000, 22, geom_photic) # Slope
geom_photic <- ifel(geom_photic == 30 & bat <= -4000, 32, geom_photic) # Rise
geom_photic <- ifel(geom_photic == 40 & bat <= -4000, 42, geom_photic) # Basin

# Isolate deep seamounts (Depth < -200m)
geom_photic <- ifel(geom_photic %in% c(50, 51) & bat < -200, 52, geom_photic)

```
### 3.4. Final Classification and Attribute Table
The raw IDs are reclassified into a sequential continuous index (1 to 9). We then embed a customized Raster Attribute Table (RAT) and color palette directly into the .tif metadata to ensure standardized visualization in QGIS or downstream applications.

```{r symbols, eval=FALSE}
# Sequential Reclassification Matrix
alpha_matrix <- matrix(c(
  11, 1,  # A1. Euphotic Shelf
  10, 2,  # A2. Mesophotic Shelf
  20, 3,  # B3. Bathyal Slope
  22, 3,  # Merged B3
  30, 4,  # C3. Bathyal Continental Rise
  32, 5,  # C4. Abyssal Continental Rise
  40, 6,  # D4. Abyssal Oceanic Basin
  42, 6,  # Merged D4
  51, 7,  # E1. Euphotic Seamount
  50, 8,  # E2. Mesophotic Seamount
  52, 9   # E3. Bathyal Seamount
), ncol = 2, byrow = TRUE)

geom_photic_id <- classify(geom_photic, alpha_matrix)
geom_photic_id <- as.factor(geom_photic_id)

# Export the L2 classified model
writeRaster(geom_photic_id, 
            'outputs/l2_biological_zones.tif', 
            datatype = "INT1U",
            overwrite = TRUE)

```


## 4. Level 3 - Substrate and Benthic Habitat Integration

The third hierarchical level combines structural geomorphology, benthic light penetration (Level 2), and seafloor substrate composition to yield the continuous benthic habitat map. Substrate characteristics are derived from an adapted EUNIS sediment classification model (Long, 2006) merged with biogenic reef formations. To avoid propagating uncertainty from sparse open-ocean sampling, substrate modeling is constrained exclusively to photic shelf domains and shallow seamount peaks.

### 4.1. Substrate Classification Framework (Modified EUNIS Folk)

Sediment textures are classified from normalized gravel, sand, and mud fractions into four primary textural classes, supplemented by an independent biogenic substrate class:

1. **Coarse sediment:** Defined where gravel is dominant or forms significant gravelly sand mixtures:
   $$\text{Gravel} \ge 80\% \quad \lor \quad \left(\text{Gravel} \ge 5\% \quad \land \quad \frac{\text{Sand}}{\text{Sand} + \text{Mud}} \ge 90\%\right)$$
2. **Mixed sediment:** Heterogeneous mixtures with intermediate gravel content:
   $$5\% \le \text{Gravel} < 80\% \quad \land \quad \frac{\text{Sand}}{\text{Sand} + \text{Mud}} < 90\%$$
3. **Sand:** Sand-dominated benthos with negligible gravel:
   $$\text{Gravel} < 5\% \quad \land \quad \frac{\text{Sand}}{\text{Sand} + \text{Mud}} \ge 90\%$$
4. **Mud to muddy sand:** Silt- and clay-dominated benthos:
   $$\text{Gravel} < 5\% \quad \land \quad \frac{\text{Sand}}{\text{Sand} + \text{Mud}} < 90\%$$
5. **Biogenic substrate:** High-priority structural formations (e.g., rhodolith beds, biogenic banks) rasterized directly from surveyed vector geometries.

```{r substrate_alignment, eval=FALSE}
library(terra)
library(dplyr)
library(scales)

# Load Level 2 biological zones and the EUNIS 5-class substrate raster
geom_photic <- rast('outputs/l2_biological_zones.tif')
sediments   <- rast('data/processed/sediment_eunis_classification.tif')

# Align sediment raster to reference photic grid using nearest neighbor
# Nearest neighbor preserves discrete integer class IDs without numerical smoothing
sediments_aligned <- project(sediments, geom_photic, method = 'near')
```

### 4.2. Spatial Map Algebra (L2 x Substrate Intersection)

The operational intersection assigns detailed sedimentary subdivisions (IDs 1 to 5) to the shallow and photic structural zones, while deep-sea domains retain their macro-geomorphological identities:

* **Subdivided zones:** Euphotic Shelf (`1`), Mesophotic Shelf (`2`), Euphotic Seamount (`7`), Mesophotic Seamount (`8`).
* **Deep macro-zones:** Bathyal Slope (`3`), Bathyal Rise (`4`), Abyssal Rise (`5`), Abyssal Basin (`6`), Bathyal Seamount (`9`).

```{r map_algebra_intersection, eval=FALSE}
# Execute conditional map algebra
# Multiply zone IDs by 100 to generate distinct 3-digit identifiers (e.g., 101, 102... 805)
habitats_raw <- ifel(
  geom_photic %in% c(1, 2, 7, 8),
  (geom_photic * 100) + sediments_aligned,
  geom_photic * 100
)
```

### 4.3. Mathematical Color Blending and Attribute Dictionary

To ensure that composite classes preserve visual hierarchy and cartographic consistency, colors are synthesized by blending the hexadecimal hue of the parent geomorphological zone ($C_{\text{geo}}$) with the sediment texture color ($C_{\text{sed}}$) via linear RGB interpolation:

$$C_{\text{blended}} = \text{round}\left(C_{\text{geo}} \times (1 - w) + C_{\text{sed}} \times w\right)$$

where a mixing weight of $w = 0.60$ gives slight dominance to the sedimentary texture while retaining the distinctive tonal identity of the bathymetric province.

```{r master_dictionary_palette, eval=FALSE}
# Define structural zones accepting substrate data
base_zones <- data.frame(
  Zone_ID   = c(1, 2, 7, 8),
  Prefix    = c("A1", "A2", "E1", "E2"),
  Zone_Name = c("Euphotic Shelf", "Mesophotic Shelf", "Euphotic Seamount", "Mesophotic Seamount"),
  Geo_Hex   = c("#00FFFF", "#00688B", "#FF4500", "#8B0000"),
  stringsAsFactors = FALSE
)

# Define EUNIS substrate classes and color palette
folk_classes <- data.frame(
  Sed_ID   = 1:5,
  Letter   = letters[1:5],
  Sed_Name = c("Coarse", "Mixed", "Sand", "Mud", "Biogenic"),
  Sed_Hex  = c("#B95246", "#C68F8A", "#F6D173", "#7F9953", "#FF69B4"),
  stringsAsFactors = FALSE
)

# Linear RGB color blending function
mix_colors <- function(c1, c2, weight = 0.6) {
  rgb1 <- col2rgb(c1)
  rgb2 <- col2rgb(c2)
  blended <- round(rgb1 * (1 - weight) + rgb2 * weight)
  rgb(blended[1], blended[2], blended[3], maxColorValue = 255)
}

# Generate theoretical attribute combinations for shallow environments
shallow_table <- expand.grid(Zone_ID = c(1, 2, 7, 8), Sed_ID = 1:5) %>%
  left_join(base_zones, by = "Zone_ID") %>%
  left_join(folk_classes, by = "Sed_ID") %>%
  rowwise() %>%
  mutate(
    Raw_ID       = (Zone_ID * 100) + Sed_ID,
    Habitat_Name = paste0(Prefix, Letter, ". ", Sed_Name, " ", Zone_Name),
    color        = mix_colors(Geo_Hex, Sed_Hex, weight = 0.6)
  ) %>%
  ungroup() %>%
  select(Raw_ID, Habitat_Name, color)

# Define standalone deep-water domains without substrate subdivision
deep_table <- data.frame(
  Raw_ID       = c(300, 400, 500, 600, 900),
  Habitat_Name = c(
    "B3. Bathyal Slope",
    "C3. Bathyal Continental Rise",
    "C4. Abyssal Continental Rise",
    "D4. Abyssal Oceanic Basin",
    "E3. Bathyal Seamount"
  ),
  color        = c("#FFA500", "#FFD700", "#B8860B", "#000050", "#8C3100"),
  stringsAsFactors = FALSE
)

# Merge into a single master reference dictionary
master_table <- bind_rows(shallow_table, deep_table) %>%
  arrange(Raw_ID)
```

### 4.4. Optimization and Native 8-bit GeoTIFF Export

The raw map algebra values produce sparse, multi-digit codes ($101$ to $900$) that exceed default byte ranges. To maximize software compatibility, reduce disk footprints, and embed visualization styles directly into the raster:

1. Active raster cell values are extracted using frequency statistics.
2. The dictionary is subsetted to observed classes and indexed sequentially from $1$ to $N$.
3. The raster is reclassified to contiguous 8-bit unsigned integers (`INT1U`).
4. Both the **Raster Attribute Table (RAT)** and the **Color Look-Up Table (CLUT)** are embedded directly into the GeoTIFF container.

```{r rat_clut_export, eval=FALSE}
# Identify active cell values present in the spatial domain
present_ids <- freq(habitats_raw)$value

# Filter dictionary and assign sequential 8-bit integer IDs (1 to N)
final_rat <- master_table %>%
  filter(Raw_ID %in% present_ids) %>%
  mutate(ID_8bit = row_number())

# Reclassify raw raster to compact 8-bit sequence
reclass_matrix <- as.matrix(final_rat[, c("Raw_ID", "ID_8bit")])
habitats_8bit  <- classify(habitats_raw, reclass_matrix, others = NA)

# Convert SpatRaster to categorical factor
habitats_8bit <- as.factor(habitats_8bit)

# Embed Raster Attribute Table (RAT)
levels(habitats_8bit) <- data.frame(
  ID      = final_rat$ID_8bit,
  Habitat = final_rat$Habitat_Name
)

# Embed Color Look-Up Table (CLUT)
coltab(habitats_8bit) <- data.frame(
  value = final_rat$ID_8bit,
  color = final_rat$color
)

# Export optimized 8-bit unsigned GeoTIFF
writeRaster(
  habitats_8bit,
  filename  = "outputs/l3_benthic_habitats_v1.tif",
  datatype  = "INT1U",
  overwrite = TRUE
)

# Export accompanying CSV legend
write.csv(final_rat, "outputs/l3_benthic_habitats_legend.csv", row.names = FALSE)
```