---
title: "REVIMAR-MAP: Benthic Marine Habitats Mapping"
author: "Tiago Gandra"
date: today
format:
  html:
    toc: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Project Overview

The **REVIMAR-MAP** project aims to provide a cartographic and methodological baseline for the predictive classification of benthic marine habitats on the Brazilian continental margin. This open-science workflow generates wall-to-wall spatial models to support Marine Spatial Planning (MSP) and environmental assessments, aligning with the guidelines of the Ministry of Environment and Climate Change (MMA) and actions funded by FUNBIO.

## 3. Geospatial Processing Pipeline (Level 1 - Oceanic Provinces)

The methodology adopts a multiscalar top-down approach. The first level of the classification hierarchy (Level 1) partitions the marine environment into broad geomorphological provinces based on bathymetry, slope, and major structural features.

### 3.1. Data Preparation and Bathymetric Smoothing

All spatial data is standardized to the Albers Equal Area Conic projection (Brazil). The raw bathymetric data (e.g., GEBCO) is smoothed using a 7x7 focal mean filter to remove high-frequency noise and local artifacts before deriving morphometric variables.

```{r smoothing, eval=FALSE}
# Define Albers projection for Brazil
crs_albers_brasil <- "+proj=aea +lat_0=-12 +lon_0=-54 +lat_1=-2 +lat_2=-22 +x_0=5000000 +y_0=10000000 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"

# Load and project bathymetry
bat_gebco <- rast('data/raw/gebco/gebco_raw.tif')
bat <- project(bat_gebco, crs_albers_brasil)

# Clamp values to isolate marine areas (depths <= 0)
bat <- clamp(bat, upper = 0)

# Apply a 7x7 focal mean filter to smooth the terrain
bat <- focal(bat, w = 7, fun = "mean", na.rm = TRUE)
```

### 3.2. Slope Calculation

Slope is essential to distinguish flat plains (basins and shelves) from steep structural zones (slopes). A focal maximum filter is also applied to emphasize the steepest gradients.

```{r slope_calculation, eval=FALSE}
# Calculate slope in degrees
slope_deg <- terrain(bat, v = "slope", unit = "degrees")

# Apply focal max to highlight structural breaks
slope_degf <- focal(slope_deg, w = 7, fun = 'max', na.rm = TRUE)
```

### 3.3. Sequential Map Algebra (Province Classification)

We apply strict depth and slope thresholds using sequential map algebra to delineate the oceanic provinces.

```{r map_algebra, eval=FALSE}
# Create an empty raster template based on bathymetry
benthic_zones <- rast(bat)
values(benthic_zones) <- NA

# Class 4: Oceanic Basin (Abyssal Plains)
# Everything deeper than 5000m, regardless of minor slope variations
benthic_zones[is.na(benthic_zones) & bat <= -5000] <- 4

# Class 1: Continental Shelf
# Shallower than the shelf break (-250m) with gentle slopes, or strictly shallower than -90m
benthic_zones[bat >= -250 & slope_deg < 0.5] <- 1
benthic_zones[bat >= -90] <- 1

# Class 2: Continental Slope
# Intermediate depths with a steep gradient
benthic_zones[is.na(benthic_zones) & bat >= -2500] <- 2
benthic_zones[is.na(benthic_zones) & slope_deg >= 0.5 & bat >= -3500] <- 2

# Class 3: Continental Rise
# Remaining unclassified intermediate to deep waters with a gentle gradient
benthic_zones[is.na(benthic_zones) & slope_deg >= 0.1] <- 3
benthic_zones[is.na(benthic_zones) & bat <= -3500] <- 4
```

### 3.4. Seamounts Integration

Isolated topographic peaks (seamounts) mapped independently in QGIS are integrated into the classification as a distinct Level 1 feature.

```{r seamounts, eval=FALSE}
# Load seamounts vector and rasterize
seamounts_vec <- read_sf('data/processed/shapes/gis_revimar.gpkg', layer = 'seamounts')
seamounts_vec <- st_transform(seamounts_vec, crs_albers_brasil)

seamounts_rast <- rasterize(seamounts_vec, benthic_zones, field = 5, background = NA)

# Overlay seamounts onto the benthic zones
benthic_zones <- cover(seamounts_rast, benthic_zones)
```

### 3.5. Noise Reduction and Polishing

To create a cartographically sound product, we apply noise reduction algorithms. A focal modal filter smooths boundaries, and a sieve algorithm removes isolated pixels (salt-and-pepper effect). Seamounts are reapplied at the end to ensure small, valid features are not deleted by the sieve process.

```{r noise_reduction, eval=FALSE}
# 1. Apply focal modal filter for boundary smoothing
focal_window <- matrix(1, nrow = 21, ncol = 21)
benthic_zones_m <- focal(benthic_zones, w = focal_window, fun = 'modal', na.rm = TRUE)

# 2. Sieve: Remove small isolated polygon fragments (< 5000 pixels)
pixels_tresholds <- 5000
benthic_zones_c <- terra::sieve(benthic_zones_m, threshold = pixels_tresholds, directions = 8)

# 3. Restore seamounts to prevent accidental deletion by the sieve
benthic_zones_final <- cover(seamounts_rast, benthic_zones_c)
```
