---
title: "REVIMAR-MAP: Benthic Marine Habitats Mapping"
author: "Tiago Gandra"
date: today
format:
  html:
    toc: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Project Overview

The **REVIMAR-MAP** project aims to provide a cartographic and methodological baseline for the predictive classification of benthic marine habitats on the Brazilian continental margin. This open-science workflow generates wall-to-wall spatial models to support Marine Spatial Planning (MSP) and environmental assessments, aligning with the guidelines of the Ministry of Environment and Climate Change (MMA) and actions funded by FUNBIO.

## 2. Environment Setup

Load the required geospatial modeling and metadata management libraries.

```{r libraries}
library(terra)
library(zen4R)
library(dplyr)
```

## 3. Geospatial Processing Pipeline

The methodology adopts a multiscalar top-down approach. We use physical variables such as geomorphology, photic zones, and sedimentology as proxies for biodiversity distribution.

### 3.1. Bathymetric Smoothing

Raw bathymetric data (e.g., GEBCO) is smoothed using a 7x7 focal mean filter to remove high-frequency noise and local artifacts before deriving morphometric variables.

```{r smoothing, eval=FALSE}
# Load raw bathymetry raster
bathy_raw <- rast("input_data/gebco_raw.tif")

# Define a 7x7 weight matrix for the focal mean
focal_weights <- matrix(1, nrow = 7, ncol = 7)

# Apply the focal filter
bathy_smoothed <- focal(bathy_raw, w = focal_weights, fun = mean, na.rm = TRUE)
```

### 3.2. Slope and Macro-relief Compartmentation

Strict slope thresholds partition the macro relief, establishing boundaries between distinct geomorphological zones. For instance, separating the Continental Rise (0.1° to 1.2°) from the Abyssal Plains (\< 0.1°).

```{r slope_calculation, eval=FALSE}
# Calculate slope in degrees
slope_deg <- terrain(bathy_smoothed, v = "slope", unit = "degrees")

# Reclassify based on established thresholds
# Reclassification matrix format: [lower_bound, upper_bound, class_id]
reclass_matrix <- matrix(c(
  0.0, 0.1, 1,  # 1 = Abyssal Plains
  0.1, 1.2, 2,  # 2 = Continental Rise
  1.2, 90,  3   # 3 = Slope / Other topographic features
), ncol = 3, byrow = TRUE)

macro_relief <- classify(slope_deg, reclass_matrix)
```

### 3.3. Bathymetric Position Index (BPI)

The BPI delineates critical retention or exposure topographic features, such as valleys, canyons, and seamounts, which are essential proxies for benthic habitats.

```{r bpi, eval=FALSE}
# Example: Broad-scale BPI using a larger focal neighborhood
scale_factor <- 25
bpi_broad <- bathy_smoothed - focal(bathy_smoothed, w = matrix(1, scale_factor, scale_factor), fun = mean)
```

## 4. Metadata and Data Archiving (Zenodo & GitHub Integration)

The resulting geospatial datasets are archived on Zenodo to ensure long-term custody, generating a permanent DOI that links back to the official GitHub repository for algorithm version control.

```{r zenodo_archiving, eval=FALSE}
# Retrieve Zenodo token and initialize manager
zenodo_token <- Sys.getenv("ZENODO_TOKEN")
zenodo <- ZenodoManager$new(token = zenodo_token)

# Retrieve the official record (e.g., creating a new version)
record_id <- "22834505"
my_deposit <- zenodo$depositRecordVersion(record_id)

# Update metadata for the new release
my_deposit$setVersion("0.5")
my_deposit$setLicense("cc-by-4.0")
my_deposit$setKeywords(c("R", "Marine Habitats", "Benthic", "Active Development"))

# Link to the GitHub repository containing this script
my_deposit$addRelatedIdentifier(
  relation = "isSupplementedBy", 
  identifier = "[https://github.com/mma-brasil/revimar-map](https://github.com/mma-brasil/revimar-map)",
  scheme = "url"
)

# Synchronize metadata with Zenodo (Publishing step requires explicit confirmation)
my_deposit <- zenodo$depositRecord(my_deposit)
# zenodo$publishRecord(my_deposit$id)
```
